---
title: "Study the presence of FHB QTLs in the URN/URSN population"
author: Charlotte Brault
output: html_document
date: '`r format(Sys.time(), "%d %B, %Y")`'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Libraries

```{r load libraries}
suppressPackageStartupMessages(library(here)) # root directory
suppressPackageStartupMessages(library(data.table)) # load data
suppressPackageStartupMessages(library(apercu)) # ap
suppressPackageStartupMessages(library(dplyr)) # formatting
suppressPackageStartupMessages(library(ggplot2)) # plot
suppressPackageStartupMessages(library(paletteer)) # plot palette
suppressPackageStartupMessages(library(janitor)) # formatting
suppressPackageStartupMessages(library(vcfR)) # handle vcf
suppressPackageStartupMessages(library(forcats)) # reorder factor 
suppressPackageStartupMessages(library(plotly)) # plot interactive
suppressPackageStartupMessages(library(ggpubr)) #nice plot   
suppressPackageStartupMessages(library(cowplot)) # combine plots   
suppressPackageStartupMessages(library(ggbeeswarm)) # nice plot  
suppressPackageStartupMessages(library(bigstatsr)) # nice plot  
suppressPackageStartupMessages(library(ggrepel)) # plot text  
suppressPackageStartupMessages(library(CGPfunctions)) # two-way anova  
suppressPackageStartupMessages(library(easystats)) # anova eta squared  
suppressPackageStartupMessages(library(multcomp)) # anova multiple comparison  
suppressPackageStartupMessages(library(emmeans)) # marginal means  
suppressPackageStartupMessages(library(patchwork)) # combine plots  

```


# Set up 

```{r current dir}
## set up root directory: indicate the local path from the project root 
here::i_am("analysis/explore_FHB_QTL_effect_URN-URSN.Rmd")
## load custom functions, for example, 'format_curate_vcf'
source(here("code","useful_functions.R")) 
pop_code <- "URN_URSN"
```


This R chunk is used to assess how much time it takes to execute the R code in this document until the end:

```{r time_0}
t0 <- proc.time()
```

# Load data

Load QTL information from KASP markers, generated in `URN_URSN_QTL_frequency.Rmd`.

```{r}
p2f <- here("output",paste0("URN_URSN_geno_FHB_QTLs_combined.tsv"))
file.mtime(p2f)
df.qtls <- fread(p2f,data.table=F)
head(df.qtls)
dim(df.qtls)
table(df.qtls$QTL.name)
df.qtls$QTL.name <- recode(df.qtls$QTL.name,"Fhb5AS(PI277012)"="Fhb5AS")
```



Load genotype information and genotypic values

```{r}
p2f <- here("data","genotype_information_URN_URSN_URNFHB.tsv")
file.mtime(p2f)
geno.info <- fread(p2f, data.table = FALSE)
head(geno.info)

```

* URN population


```{r}
dat.urn <- fread(here("output","URN_geno_blues_lme4-values.tsv"))
head(dat.urn)
summary(dat.urn$HD)
dim(dat.urn) ## 1034 x 8
```



* URSN population

```{r}
dat.fhb <- fread(here("output","URN_URSN_FHB_geno_blues_lme4-values.tsv"), data.table=F)
dim(dat.fhb) # 1382 x 7
head(dat.fhb)
length(unique(dat.fhb$GID))
# ## remove heading date trait because already present in URN and URSN data
# dat.fhb$HD <- NULL
```


* Combine URN and URSN data

```{r}
dat <- merge(dat.urn, dat.fhb, by="GID", all=TRUE)
## combine heading date trait because already present in URN and URSN data
dat$HD <- coalesce(dat$HD.x, dat$HD.y) 
dat <- dat |> dplyr::select(-HD.x, -HD.y)
print_table(head(dat, n=100))
traits <- colnames(dat)[-1]
#traits <- c("DIS","VSK","DON","YLD","TWT","PROT")
```


# Format data

```{r}
## add genotypic values to QTL information
gv.qtl <- merge(df.qtls, dat, by="GID",  all.x=T, all.y=F)

## add genotype information to QTL information
gv.qtl <- merge(gv.qtl, 
                dplyr::select(geno.info, -ISCK, -SET),#[,c("GID",setdiff(colnames(geno.info),colnames(df.qtls)))],
                by="GID", all.x=T, all.y=F)

#cb <- cbind(gv.qtl$FIRST_YEAR.x, gv.qtl$FIRST_YEAR.y)
#gv.qtl$FIRST_YEAR <- coalesce(gv.qtl$FIRST_YEAR.x, gv.qtl$FIRST_YEAR.y)
gv.qtl$FIRST_YEAR <- dplyr::coalesce(gv.qtl$FIRST_YEAR.y, gv.qtl$FIRST_YEAR.x)
gv.qtl$SOURCE <- dplyr::coalesce(gv.qtl$SOURCE.x, gv.qtl$SOURCE.y)
gv.qtl$NURSERY <- dplyr::coalesce(gv.qtl$NURSERY.x, gv.qtl$NURSERY.y)

#gv.qtl$SOURCE[gv.qtl$SOURCE %in% c("OTH",NA)] <- "OTHER"

gv.qtl <- gv.qtl |> dplyr::select(-FIRST_YEAR.x, -FIRST_YEAR.y,
                                  -SOURCE.x, -SOURCE.y,
                                  -NURSERY.x, -NURSERY.y)
print_table(head(gv.qtl))

## put traits at the end of the table
gv.qtl <- gv.qtl |> relocate(all_of(traits), .after="FIRST_YEAR")
```




# Study individual QTL effects


## Fhb1 effect on FHB by source

```{r Fhb1_effect_source_interaction}
FHB.traits <- c("DIS","VSK","DON")
BPs <- c("AAFC-CA","MN","ND","SD","MT")

pal <- c("AAFC-CA"="#F5586A",MN="#ffcc33",ND="#2E8B57",SD="#3A5FCD", MT="#bd9b04")

for(tr in FHB.traits){
  dat.fit <- gv.qtl |>
    dplyr::select(all_of(c("GID","FIRST_YEAR","SOURCE", "QTL.name", "allele", tr))) |>
    filter(!is.na(.data[[tr]]),FIRST_YEAR >= 2001,
           QTL.name %in% "Fhb1",
           allele %in% c("Res","Sus"),SOURCE %in% BPs) |> 
    mutate(allele=as.factor(allele),SOURCE=as.factor(SOURCE))
  
  cat(tr)
  fit <- lm(paste0(tr, "~ FIRST_YEAR + allele+ SOURCE"), data=dat.fit)
  print(summary(fit))
  tabyl(dat.fit, SOURCE, allele)
  
  ## two-way ANOVA
  
  fct <- Plot2WayANOVA(formula =as.formula(paste0(tr,"~ allele*SOURCE")),
                       dataframe=dat.fit, errorbar.display = "CI",
                       mean.color = "white",mean.label = TRUE,
                       ggplot.component =
                         list(scale_color_manual(values=pal),
                              scale_fill_manual(values=pal)))
  ## remove the effect of first year beforehand?
  pos <- position_dodge(width = 0.05)
  
  plot(ggplot(fct$MeansTable, aes(x=SOURCE, y=TheMean, color=allele,
                                  ymin=LowerBoundCI,ymax=UpperBoundCI, group=allele))+
         geom_point(size=3,position=pos) +
         geom_errorbar(width=0.1, position=pos)+
         ggtitle(paste0("Estimate of Fhb1 allelic effect on ",tr))+
         ylab("Fhb1 effect") + xlab("Breeding program")+
         geom_line(position=pos, linewidth=0.8, linetype="dotted") +
         theme_bigstatsr(size.rel=0.8)+ 
         scale_color_manual(values=c("Res"="#0D803B","Sus"="#AD151A","Unk"="grey60")))
  print_table(fct$MeansTable)
}

```

## Fhb1 effect on 6 traits

```{r study_fhb1_effect_6traits}
traits.fhb <- c("DIS","VSK","DON")
plist <- list()
traits.sel <- c(traits.fhb, "YLD","PROT","TWT")
plt.arr <- list(nudgey=c(DIS=2,VSK=3,DON=2,YLD=-300,PROT=-0.8,TWT=-2),
                yeta=c(DIS=53,VSK=83,DON=55,YLD=4450,PROT=15.7,TWT=82))
# ysignif=c(DIS=65,VSK=91,DON=47,YLD=4700,PROT=16.1,TWT=84),
# ymax=c(DIS=74,VSK=103,DON=55,YLD=5200,PROT=15.7,TWT=86.5))
marginal.means <- list()
for(tr in traits.sel){
  dat2plot <- gv.qtl |>
    dplyr::select(all_of(c("GID","QTL.name","FIRST_YEAR","SOURCE", "allele", tr))) |> 
    filter(!is.na(.data[[tr]]),
           QTL.name == "Fhb1",
           FIRST_YEAR >= 1995,
           allele %in% c("Res","Sus")) |>
    mutate(allele=factor(allele, levels=c("Res","Sus"),
                         labels=c("Resistant","Susceptible")))
  #dat2plot$FIRST_YEAR[dat2plot$FIRST_YEAR < 1968] <- 1968
  cat(tr)
  ## Estimate the effect of Fhb1 on the trait while controlling for the first year
  fit <- lm(paste0(tr, "~ FIRST_YEAR + allele"), data=dat2plot)
  print(summary(fit))
  # wilcox.test(x=dat2plot[dat2plot$allele == "Resistant",tr],
  #             y=dat2plot[dat2plot$allele == "Susceptible",tr])
  pval <- anova(fit)[["Pr(>F)"]][2]
  emm <- as.data.frame(emmeans::emmeans(fit, specs = "allele") )
  marginal.means[[tr]] <- emm
  eta2 <- effectsize::eta_squared(fit)
  
  ## add mean value in the two categories
  #summ.text <- dat2plot |> group_by(allele) |> summarize(mean=mean(.data[[tr]]))
  
  ## add number of obs
  xlabs <- paste(levels(dat2plot$allele),"\n(N=",table(dat2plot$allele),")",sep="")
  
  plist[[tr]] <- ggplot(dat2plot, aes(x=allele, y=.data[[tr]]))+
    geom_quasirandom(alpha=0.3, width=0.25, size=0.8, aes(color=FIRST_YEAR))+
    geom_boxplot(fill=NA,lwd=1, show.legend=F, outliers = FALSE, width=0.25, color="maroon")+
    xlab(paste0("Fhb1 allele"))+
    ## expand the y axis with mult
    scale_y_continuous(expand = expansion(mult = c(0, .2)))+
    scale_x_discrete(labels=xlabs)+
    guides(color=guide_colorbar(barwidth=12 ))+
    geom_point(data=emm, aes(x=allele, y=emmean), 
               size=3, shape=4, color="black",stroke=1.2,)+
    geom_errorbar(data=emm, aes(x=allele, y=emmean, ymin=lower.CL, ymax=upper.CL), 
                  width=0.01, color="black", size=0.8)+
    geom_label_repel(data=emm, aes(x=allele, y=emmean, label=round(emmean,2)), 
                     color="black", size=4, nudge_x=0.38, 
                     nudge_y=plt.arr$nudgey[tr], min.segment.length=0)+
    ggsignif::stat_signif(comparisons = list(c("Resistant","Susceptible")), 
                          #test="wilcox.test",
                          annotations=sprintf("p = %.2g", pval),
                          vjust=-0.5)+
    annotate("text",label=paste0("\u03b7\u00b2 = ",round(eta2$Eta2_partial[2],4)),
             x=1.5, y=plt.arr$yeta[tr],size=4, color="black")+
    scale_color_viridis_c( name="First year")+
    theme_bigstatsr(size.rel=0.75) +
    theme(legend.position="bottom", legend.direction="horizontal",
          legend.title.position = "top",
          axis.text.x=element_text(size=10), axis.title.x=element_blank())
  plot(plist[[tr]])
  rm(fit, dat2plot)
}
ggarrange(plotlist=plist, ncol=2, nrow=3,common.legend = TRUE, legend="bottom",labels = "AUTO")
ggsave(here("output","figures",paste0(pop_code,"_Fhb1_effects_6traits_distrib_v3.png")),
       height=8,width=11,scale=0.85)
```

Study marginal means for Fhb1 effect on traits
```{r}
marginal.means |> bind_rows(.id="trait") |> group_by(trait) |> 
  summarize(perc.diff= (emmean[allele=="Resistant"] - 
                          emmean[allele=="Susceptible"]) /
              emmean[allele=="Susceptible"] * 100)
```


## Fhb5 effect on 6 traits

```{r study_fhb5_effect_6traits}
traits.fhb <- c("DIS","VSK","DON")
plist <- list()
traits.sel <- c(traits.fhb, "YLD","PROT","TWT")
plt.arr <- list(nudgey=c(DIS=2,VSK=3,DON=2,YLD=-300,PROT=-0.8,TWT=-2),
                yeta=c(DIS=53,VSK=83,DON=55,YLD=4450,PROT=15.7,TWT=82))
# ysignif=c(DIS=65,VSK=91,DON=47,YLD=4700,PROT=16.1,TWT=84),
# ymax=c(DIS=74,VSK=103,DON=55,YLD=5200,PROT=15.7,TWT=86.5))
marginal.means <- list()
for(tr in traits.sel){
  dat2plot <- gv.qtl |>
    dplyr::select(all_of(c("GID","QTL.name","FIRST_YEAR","SOURCE", "allele", tr))) |> 
    filter(!is.na(.data[[tr]]),
           QTL.name == "Fhb5",
           FIRST_YEAR >= 1995,
           allele %in% c("Res","Sus")) |>
    mutate(allele=factor(allele, levels=c("Res","Sus"),
                         labels=c("Resistant","Susceptible")))
  #dat2plot$FIRST_YEAR[dat2plot$FIRST_YEAR < 1968] <- 1968
  cat(tr)
  ## Estimate the effect of Fhb1 on the trait while controlling for the first year
  fit <- lm(paste0(tr, "~ FIRST_YEAR + allele"), data=dat2plot)
  print(summary(fit))
  pval <- anova(fit)[["Pr(>F)"]][2]
  emm <- as.data.frame(emmeans::emmeans(fit, specs = "allele") )
  marginal.means[[tr]] <- emm
  eta2 <- eta_squared(fit)
  
  ## add mean value in the two categories
  #summ.text <- dat2plot |> group_by(allele) |> summarize(mean=mean(.data[[tr]]))
  
  ## add number of obs
  xlabs <- paste(levels(dat2plot$allele),"\n(N=",table(dat2plot$allele),")",sep="")
  
  plist[[tr]] <- ggplot(dat2plot, aes(x=allele, y=.data[[tr]], color=FIRST_YEAR))+
    geom_quasirandom(alpha=0.3, width=0.25, size=0.8)+
    
    geom_boxplot(fill=NA,lwd=1, show.legend=F, outliers = FALSE, width=0.25, color="maroon")+
    #stat_summary(fun=mean, geom="point",position=position_dodge(0.85),shape=4,size=4, stroke=1.2, color="black")+
    xlab(paste0("Fhb5 allele"))+
    ## expand the y axis with mult
    scale_y_continuous(expand = expansion(mult = c(0, .2)))+
    #scale_y_continuous(limits=c(min(dat2plot[[tr]], na.rm=T), plt.arr$ymax[tr]))+
    scale_x_discrete(labels=xlabs)+
    guides(color=guide_colorbar(barwidth=12 ))+
    geom_point(data=emm, aes(x=allele, y=emmean), 
               size=3, shape=4, color="black",stroke=1.2,)+
    geom_label_repel(data=emm, aes(x=allele, y=emmean, label=round(emmean,2)), 
                     color="black", size=4, nudge_x=0.38, 
                     nudge_y=plt.arr$nudgey[tr], min.segment.length=0)+
    #stat_compare_means(comparisons = list(c("Res","Sus")))+ # Add significance levels
    ggsignif::stat_signif(comparisons = list(c("Resistant","Susceptible")), 
                          #test="wilcox.test",
                          annotations=sprintf("p = %.2g", pval),
                          #y_position=plt.arr$ysignif[tr], 
                          vjust=-0.5)+
    annotate("text",label=paste0("\u03b7\u00b2 = ",round(eta2$Eta2_partial[2],4)),
             x=1.5, y=plt.arr$yeta[tr],size=4, color="black")+
    scale_color_viridis_c( name="First year")+
    theme_bigstatsr(size.rel=0.75) +
    theme(legend.position="bottom", legend.direction="horizontal",
          legend.title.position = "top",
          axis.text.x=element_text(size=10), axis.title.x=element_blank())
  plot(plist[[tr]])
  rm(fit, dat2plot)
}
ggarrange(plotlist=plist, ncol=2, nrow=3,common.legend = TRUE, legend="bottom",labels = "AUTO")
ggsave(here("output","figures",paste0(pop_code,"_Fhb5_effects_6traits_distrib_v3.png")),
       height=8,width=11,scale=0.85)
```


Study marginal means for Fhb5 effect on traits
```{r}
marginal.means |> bind_rows(.id="trait") |> group_by(trait) |> 
  summarize(perc.diff= (emmean[allele=="Resistant"] -
                          emmean[allele=="Susceptible"]) /
              emmean[allele=="Susceptible"] * 100)
```

## Fhb5AS effect on 6 traits

```{r study_fhb5as_effect_6traits}
traits.fhb <- c("DIS","VSK","DON")
plist <- list()
traits.sel <- c(traits.fhb, "YLD","PROT","TWT")
plt.arr <- list(nudgey=c(DIS=2,VSK=3,DON=2,YLD=-300,PROT=-0.8,TWT=-2),
                yeta=c(DIS=53,VSK=83,DON=55,YLD=4450,PROT=15.7,TWT=82))
# ysignif=c(DIS=65,VSK=91,DON=47,YLD=4700,PROT=16.1,TWT=84),
# ymax=c(DIS=74,VSK=103,DON=55,YLD=5200,PROT=15.7,TWT=86.5))
marginal.means <- list()

for(tr in traits.sel){
  dat2plot <- gv.qtl |>
    dplyr::select(all_of(c("GID","QTL.name","FIRST_YEAR","SOURCE", "allele", tr))) |> 
    filter(!is.na(.data[[tr]]),
           QTL.name == "Fhb5AS",
           FIRST_YEAR >= 1975,
           allele %in% c("Res","Sus")) |>
    mutate(allele=factor(allele, levels=c("Res","Sus"),
                         labels=c("Resistant","Susceptible")))
  #dat2plot$FIRST_YEAR[dat2plot$FIRST_YEAR < 1968] <- 1968
  cat(tr)
  ## Estimate the effect of Fhb1 on the trait while controlling for the first year
  fit <- lm(paste0(tr, "~ FIRST_YEAR + allele"), data=dat2plot)
  print(summary(fit))
  pval <- anova(fit)[["Pr(>F)"]][2]
  emm <- as.data.frame(emmeans::emmeans(fit, specs = "allele") )
  marginal.means[[tr]] <- emm
  eta2 <- eta_squared(fit)
  
  ## add mean value in the two categories
  #summ.text <- dat2plot |> group_by(allele) |> summarize(mean=mean(.data[[tr]]))
  
  ## add number of obs
  xlabs <- paste(levels(dat2plot$allele),"\n(N=",table(dat2plot$allele),")",sep="")
  
  plist[[tr]] <- ggplot(dat2plot, aes(x=allele, y=.data[[tr]], color=FIRST_YEAR))+
    geom_quasirandom(alpha=0.3, width=0.25, size=0.8)+
    
    geom_boxplot(fill=NA,lwd=1, show.legend=F, outliers = FALSE, width=0.25, color="maroon")+
    #stat_summary(fun=mean, geom="point",position=position_dodge(0.85),shape=4,size=4, stroke=1.2, color="black")+
    xlab(paste0("Fhb5AS allele"))+
    ## expand the y axis with mult
    scale_y_continuous(expand = expansion(mult = c(0, .2)))+
    #scale_y_continuous(limits=c(min(dat2plot[[tr]], na.rm=T), plt.arr$ymax[tr]))+
    scale_x_discrete(labels=xlabs)+
    guides(color=guide_colorbar(barwidth=12 ))+
    geom_point(data=emm, aes(x=allele, y=emmean), 
               size=3, shape=4, color="black",stroke=1.2,)+
    geom_label_repel(data=emm, aes(x=allele, y=emmean, label=round(emmean,2)), 
                     color="black", size=4, nudge_x=0.38, 
                     nudge_y=plt.arr$nudgey[tr], min.segment.length=0)+
    #stat_compare_means(comparisons = list(c("Res","Sus")))+ # Add significance levels
    ggsignif::stat_signif(comparisons = list(c("Resistant","Susceptible")), 
                          #test="wilcox.test",
                          annotations=sprintf("p = %.2g", pval),
                          #y_position=plt.arr$ysignif[tr], 
                          vjust=-0.5)+
    annotate("text",label=paste0("\u03b7\u00b2 = ",round(eta2$Eta2_partial[2],4)),
             x=1.5, y=plt.arr$yeta[tr],size=4, color="black")+
    scale_color_viridis_c( name="First year")+
    theme_bigstatsr(size.rel=0.75) +
    theme(legend.position="bottom", legend.direction="horizontal",
          legend.title.position = "top",
          axis.text.x=element_text(size=10), axis.title.x=element_blank())
  plot(plist[[tr]])
  rm(fit, dat2plot)
}
ggarrange(plotlist=plist, ncol=2, nrow=3,common.legend = TRUE, legend="bottom",labels = "AUTO")
ggsave(here("output","figures",paste0(pop_code,"_Fhb5AS_effects_6traits_distrib.png")),
       height=8,width=11,scale=0.85)
```

Study marginal means for Fhb5AS effect on traits
```{r}
marginal.means |> bind_rows(.id="trait") |> group_by(trait) |> 
  summarize(perc.diff= (emmean[allele=="Resistant"] - 
                          emmean[allele=="Susceptible"]) / 
              emmean[allele=="Susceptible"] * 100)
```

# Combination of QTLs

## Model selection

```{r comb_qtl, message=FALSE, include=FALSE}
mod.sel <- list()
for(tr in traits.sel){
  dat2fit <- gv.qtl |>
    dplyr::select(all_of(c("GID","QTL.name","FIRST_YEAR","SOURCE", "allele", tr))) |> 
    filter(!is.na(.data[[tr]]),
           QTL.name %in% c("Fhb1","Fhb5","Fhb5AS"),
           FIRST_YEAR >= 1995,
           allele %in% c("Res","Sus")) |>
    tidyr::pivot_wider(names_from=QTL.name, values_from=allele) |> 
    ## replace NA by Sus
    # mutate(Fhb1=metan::replace_na(Fhb1, replacement="Sus"),
    #        Fhb5=metan::replace_na(Fhb5, replacement="Sus"),
    #        Fhb5AS=metan::replace_na(Fhb5AS, replacement="Sus"))
    mutate(Fhb1=metan::replace_na(Fhb1, replacement="Unk"),
           Fhb5=metan::replace_na(Fhb5, replacement="Unk"),
           Fhb5AS=metan::replace_na(Fhb5AS, replacement="Unk"))
  
  ## Estimate the effect of Fhb1 on the trait while controlling for the first year
  fit <- lm(paste0(tr, "~ 1+ FIRST_YEAR + Fhb5AS * Fhb1 * Fhb5"), data=dat2fit)
  fit.step <- step(fit)
  #anova(fit.step)
  #print(summary(fit.step))
  mod.sel[[tr]] <- fit.step
}
```

```{r}
lapply(mod.sel, anova)
```

```{r}
lapply(mod.sel, summary)
sapply(mod.sel, \(x) round(summary(x)$adj.r.squared,3))

```


## Interaction for DIS

Analyze interaction effects
```{r}
tr <- "DIS"
## fit two way anova
mod.sel[[tr]]
#mod.sel[[tr]]$coefficients
tkt <- TukeyHSD(aov(mod.sel[[tr]]), which = c("Fhb1","Fhb5","Fhb5AS"))#,"Fhb1:Fhb5"))
tkt.df <- do.call(rbind, list(
  cbind(effect="Fhb1", diff.effect=rownames(tkt$Fhb1), tkt$Fhb1),
  cbind(effect="Fhb5", diff.effect=rownames(tkt$Fhb5),tkt$Fhb5),
  cbind(effect="Fhb5AS", diff.effect=rownames(tkt$Fhb5AS),tkt$Fhb5AS)))
  #cbind(effect="Fhb1:Fhb5", diff.effect=rownames(tkt$`Fhb1:Fhb5`),tkt$`Fhb1:Fhb5`)))
tkt.df <- as.data.frame(tkt.df)
## remove unknown rows
tkt.df <- tkt.df[!grepl("Unk", tkt.df$diff.effect),]
tkt.df[,3:ncol(tkt.df)] <- apply(tkt.df[,3:ncol(tkt.df)],2, as.numeric)
tkt.df$yaxis <- paste0(tkt.df$effect, " - ", tkt.df$diff.effect)


tkt.df$effect <- factor(tkt.df$effect, levels=c("Fhb1","Fhb5","Fhb5AS"))#,"Fhb1:Fhb5"))
tkt.df <- arrange(tkt.df, desc(effect))
tkt.df$yaxis <- factor(tkt.df$yaxis, levels=unique(tkt.df$yaxis))

ggplot(tkt.df, aes(x=diff, y=yaxis, xmin=lwr, xmax=upr))+
  geom_point(size=3)+
  geom_errorbarh(height=0.2)+
  ## color differently interaction and main effects
  geom_point(data=tkt.df[tkt.df$effect != "Fhb1:Fhb5",], aes(x=diff), color="maroon",size=3)+
  geom_vline(xintercept = 0, linetype="dotted", color="grey50")+
  xlab("Difference in means")+
  ylab("Comparison")+
  ggtitle(paste0("Tukey HSD test for ",tr," trait"))+
  theme_bigstatsr(size.rel=0.8)+
  theme(axis.text.y = element_text(size=10))
# 

# dat2fit <- gv.qtl |>
#     dplyr::select(all_of(c("GID","QTL.name","FIRST_YEAR","SOURCE", "allele", tr))) |> 
#     filter(!is.na(.data[[tr]]),
#            QTL.name %in% c("Fhb1","Fhb5","Fhb5AS"),
#            FIRST_YEAR >= 1995,
#            allele %in% c("Res","Sus")) |>
#     tidyr::pivot_wider(names_from=QTL.name, values_from=allele)
# Plot2WayANOVA(formula = DIS ~ Fhb1 * Fhb5 , dataframe = dat2fit)
# ggcoefstats(mod.sel[[tr]],exclude.intercept=TRUE)

```

<!-- ```{r} -->
<!-- marginal = emmeans(mod.sel[[tr]], -->
<!--                    ~ FIRST_YEAR+Fhb1+Fhb5+Fhb5AS+Fhb1:Fhb5) -->

<!-- #pairs(marginal,adjust="tukey") -->

<!-- cld=cld(marginal,alpha=0.05,Letters=letters,adjust="tukey") -->
<!-- cld -->
<!-- pos <- position_dodge(0.2) -->
<!-- ggplot(cld, aes(x=Fhb1, y=emmean, ymin=lower.CL, ymax=upper.CL, -->
<!--                 color=Fhb5, group=Fhb5))+ -->
<!--   facet_wrap(~Fhb5AS, labeller=label_both)+ -->
<!--   geom_line(linewidth=1.3,alpha=0.3, position=pos)+ -->
<!--   geom_point(size=3, position=pos) + -->
<!--   scale_x_discrete(labels=c("Res"="Fhb1 Resistant","Sus"="Fhb1 Susceptible"))+ -->
<!--   scale_color_manual(values=c("Res"="#0D803B","Sus"="#AD151A"))+ -->
<!--   labs(x="Fhb1 allele", y=paste0("Marginal means for ",tr))+ -->
<!--   geom_errorbar(width=0.1, position=pos)+ -->
<!--   theme_bigstatsr(size.rel=0.8)+ -->
<!--   theme(axis.title.x=element_blank(), -->
<!--         legend.position="inside", -->
<!--         legend.position.inside=c(0.8,0.05), -->
<!--         legend.direction="horizontal", -->
<!--         legend.background = element_rect(fill="white",color="black")) -->

<!-- ggplot(cld, aes(x=Fhb5, y=emmean, ymin=lower.CL, ymax=upper.CL, -->
<!--                 color=Fhb1, group=Fhb1))+ -->
<!--   facet_wrap(~Fhb5AS, labeller=label_both)+ -->
<!--   geom_line(linewidth=1.3,alpha=0.3, position=pos)+ -->
<!--   geom_point(size=3, position=pos) + -->
<!--   scale_x_discrete(labels=c("Res"="Fhb5 Resistant","Sus"="Fhb5 Susceptible"))+ -->
<!--   scale_color_manual(values=c("Res"="#0D803B","Sus"="#AD151A"))+ -->
<!--   labs(x="Fhb1 allele", y=paste0("Marginal means for ",tr))+ -->
<!--   geom_errorbar(width=0.1, position=pos)+ -->
<!--   theme_bigstatsr(size.rel=0.8)+ -->
<!--   theme(axis.title.x=element_blank(), -->
<!--         legend.position="inside", -->
<!--         legend.position.inside=c(0.8,0.05), -->
<!--         legend.direction="horizontal", -->
<!--         legend.background = element_rect(fill="white",color="black")) -->
<!-- ggsave(here("output","figures",paste0(pop_code,"_Fhb1_Fhb5_interaction_",tr,".png")), -->
<!--        height=6,width=8,scale=0.8) -->
<!-- ``` -->
```{r}
dat2fit <- gv.qtl |>
  dplyr::select(all_of(c("GID","QTL.name","FIRST_YEAR","SOURCE", "allele", tr))) |> 
  filter(!is.na(.data[[tr]]),
         QTL.name %in% c("Fhb1","Fhb5","Fhb5AS"),
         FIRST_YEAR >= 1995,
         allele %in% c("Res","Sus"),!is.na(allele)) |>
  tidyr::pivot_wider(names_from=QTL.name, values_from=allele) 
dat2fit |> group_by(Fhb1,Fhb5,Fhb5AS) |> na.omit() |> 
  summarize(n=n(), mean=mean(.data[[tr]]), sd=sd(.data[[tr]]),.groups="drop") |> 
  arrange(Fhb1,Fhb5,Fhb5AS) |> print_table()
```



<!-- ```{r} -->
<!-- cld$group <- paste0(cld$Fhb5AS, " - ", cld$Fhb5) -->
<!-- pDIS <- ggplot(cld, aes(x=Fhb1, y=emmean, ymin=lower.CL, ymax=upper.CL,  -->
<!--                         color=Fhb5, shape=Fhb5AS, group=group))+ -->
<!--   #facet_wrap(~Fhb5AS, labeller=label_both)+ -->
<!--   geom_line(linewidth=1.3,alpha=0.3, position=pos)+ -->
<!--   geom_point(size=3, position=pos) + -->
<!--   scale_x_discrete(labels=c("Res"="Fhb1 Resistant","Sus"="Fhb1 Susceptible"))+ -->
<!--   scale_color_manual(values=c("Res"="#0D803B","Sus"="#AD151A"))+ -->
<!--   labs(x="Fhb1 allele", y=paste0("Marginal means for ",tr))+ -->
<!--   geom_errorbar(width=0.1, position=pos)+ -->
<!--   theme_bigstatsr(size.rel=0.8)+ -->
<!--   theme(axis.title.x=element_blank(), -->
<!--         legend.position="inside", -->
<!--         legend.position.inside=c(0.5,0.18), -->
<!--         legend.background=element_rect(fill="transparent"), -->
<!--         #legend.box.background = element_rect(color="black"), -->
<!--         legend.title.position = "top", -->
<!--         legend.title=element_text(size=12,hjust=0.5), -->
<!--         legend.direction="horizontal") -->
<!-- pDIS -->
<!-- ``` -->


Main effects: Fhb1 is the biggest effect compared to Fhb5 and Fhb5AS.
Significant interaction betweeb Fhb1 and Fhb5: Fhb5 Res is having more effect when this is the only resistance allele. The combination of the two resistance does not bring further reduction in FHB disease index (?).


<!-- ## Interaction for VSK -->

<!-- ```{r} -->
<!-- tr <- "VSK" -->
<!-- ## fit two way anova -->
<!-- mod.sel[[tr]] -->
<!-- #mod.sel[[tr]]$coefficients -->
<!-- tkt <- TukeyHSD(aov(mod.sel[[tr]]), which = c("Fhb1","Fhb5","Fhb5AS","Fhb5AS:Fhb1")) -->
<!-- tkt.df <- do.call(rbind, list( -->
<!--   cbind(effect="Fhb1", diff.effect=rownames(tkt$Fhb1), tkt$Fhb1), -->
<!--   cbind(effect="Fhb5", diff.effect=rownames(tkt$Fhb5),tkt$Fhb5), -->
<!--   cbind(effect="Fhb5AS", diff.effect=rownames(tkt$Fhb5AS),tkt$Fhb5AS), -->
<!--   cbind(effect="Fhb5AS:Fhb1", diff.effect=rownames(tkt$`Fhb5AS:Fhb1`),tkt$`Fhb5AS:Fhb1`))) -->
<!-- tkt.df <- as.data.frame(tkt.df) -->
<!-- tkt.df[,3:ncol(tkt.df)] <- apply(tkt.df[,3:ncol(tkt.df)],2, as.numeric) -->
<!-- tkt.df$yaxis <- paste0(tkt.df$effect, " - ", tkt.df$diff.effect) -->


<!-- tkt.df$effect <- factor(tkt.df$effect, levels=c("Fhb1","Fhb5","Fhb5AS","Fhb5AS:Fhb1")) -->
<!-- tkt.df <- arrange(tkt.df, desc(effect)) -->
<!-- tkt.df$yaxis <- factor(tkt.df$yaxis, levels=unique(tkt.df$yaxis)) -->

<!-- ggplot(tkt.df, aes(x=diff, y=yaxis, xmin=lwr, xmax=upr))+ -->
<!--   geom_point(size=3)+ -->
<!--   geom_errorbarh(height=0.2)+ -->
<!--   ## color differently interaction and main effects -->
<!--   geom_point(data=tkt.df[tkt.df$effect != "Fhb5AS:Fhb1",], aes(x=diff), color="maroon",size=3)+ -->
<!--   geom_vline(xintercept = 0, linetype="dotted", color="grey50")+ -->
<!--   xlab("Difference in means")+ -->
<!--   ylab("Comparison")+ -->
<!--   ggtitle(paste0("Tukey HSD test for ",tr," trait"))+ -->
<!--   theme_bigstatsr(size.rel=0.8)+ -->
<!--   theme(axis.text.y = element_text(size=10)) -->

<!-- # dat2fit <- gv.qtl |> -->
<!-- #     dplyr::select(all_of(c("GID","QTL.name","FIRST_YEAR","SOURCE", "allele", tr))) |>  -->
<!-- #     filter(!is.na(.data[[tr]]), -->
<!-- #            QTL.name %in% c("Fhb1","Fhb5","Fhb5AS"), -->
<!-- #            FIRST_YEAR >= 1995, -->
<!-- #            allele %in% c("Res","Sus")) |> -->
<!-- #     tidyr::pivot_wider(names_from=QTL.name, values_from=allele) -->
<!-- # Plot2WayANOVA(formula = VSK ~ Fhb1 * Fhb5AS , dataframe = dat2fit) -->
<!-- ``` -->

<!-- Main effects: Fhb1 is the biggest effect compared to Fhb5 and Fhb5AS. -->
<!-- Significant interaction betweeb Fhb1 and Fhb5AS: Fhb1 is having more effect than Fhb5AS in the interaction. -->

<!-- ```{r} -->
<!-- marginal = emmeans(mod.sel[[tr]], -->
<!--                    ~ FIRST_YEAR+Fhb1+Fhb5+Fhb5AS+Fhb1:Fhb5AS) -->

<!-- # pairs(marginal,adjust="tukey") -->

<!-- cld=cld(marginal,alpha=0.05,Letters=letters,adjust="tukey") -->
<!-- cld -->
<!-- pos <- position_dodge(0.2) -->
<!-- # ggplot(cld, aes(x=Fhb1, y=emmean, ymin=lower.CL, ymax=upper.CL, -->
<!-- #                 color=Fhb5AS, group=Fhb5AS))+ -->
<!-- #   facet_wrap(~Fhb5, labeller=label_both)+ -->
<!-- #   geom_line(linewidth=1.3,alpha=0.3, position=pos)+ -->
<!-- #   geom_point(size=3, position=pos) + -->
<!-- #   scale_x_discrete(labels=c("Res"="Fhb1 Resistant","Sus"="Fhb1 Susceptible"))+ -->
<!-- #   scale_color_manual(values=c("Res"="#0D803B","Sus"="#AD151A"))+ -->
<!-- #   labs(x="Fhb1 allele", y=paste0("Marginal means for ",tr))+ -->
<!-- #   geom_errorbar(width=0.1, position=pos)+ -->
<!-- #   theme_bigstatsr(size.rel=0.8)+ -->
<!-- #   theme(axis.title.x=element_blank(), -->
<!-- #         legend.position="inside", -->
<!-- #         legend.position.inside=c(0.8,0.2), -->
<!-- #         legend.background = element_rect(fill="white",color="black")) -->

<!-- ``` -->
<!-- ```{r} -->
<!-- pos <- position_dodge(0.2) -->

<!-- cld$group <- paste0(cld$Fhb5AS, " - ", cld$Fhb5) -->
<!-- pVSK <- ggplot(cld, aes(x=Fhb1, y=emmean, ymin=lower.CL, ymax=upper.CL,  -->
<!--                         color=Fhb5, shape=Fhb5AS, group=group))+ -->
<!--   #facet_wrap(~Fhb5AS, labeller=label_both)+ -->
<!--   geom_line(linewidth=1.3,alpha=0.3, position=pos)+ -->
<!--   geom_point(size=3, position=pos) + -->
<!--   scale_x_discrete(labels=c("Res"="Fhb1 Resistant","Sus"="Fhb1 Susceptible"))+ -->
<!--   scale_color_manual(values=c("Res"="#0D803B","Sus"="#AD151A"))+ -->
<!--   labs(x="Fhb1 allele", y=paste0("Marginal means for ",tr))+ -->
<!--   geom_errorbar(width=0.1, position=pos)+ -->
<!--   theme_bigstatsr(size.rel=0.8)+ -->
<!--   theme(axis.title.x=element_blank(), -->
<!--         legend.position="inside", -->
<!--         legend.position.inside=c(0.55,0.18), -->
<!--         legend.background=element_rect(fill="transparent"), -->
<!--         legend.title.position = "top", -->
<!--         legend.title=element_text(size=12,hjust=0.5), -->
<!--         legend.direction="horizontal") -->
<!-- plot(pVSK) -->
<!-- ``` -->

```{r}
# pDIS #+ pVSK + plot_layout(ncol=2) +plot_annotation(tag_levels="A")
# ggsave(here("output","figures",paste0(pop_code,"_Fhb1_Fhb5_Fhb5AS_interaction_DIS.png")),
#        height=8,width=10,scale=0.8)
```




# UpSet Plot


Plot Upset
```{r upset_plot,fig.height=7}
qtls.fhb <- unique(df.qtls$QTL.name)


dat.upset <- df.qtls |> distinct() |> 
  filter(allele %in% c("Res","Sus")) |> #,FIRST_YEAR >=1995) |> 
  tidyr::pivot_wider(names_from=QTL.name, values_from=allele,
                     id_cols=all_of(c("GID","FIRST_YEAR","NURSERY"))) |> 
  mutate(across(all_of(qtls.fhb), ~ ifelse(.x == "Res",1,0))) |> 
  left_join(dat, by="GID")
head(dat.upset)
ComplexUpset::upset(dat.upset, intersect=qtls.fhb,min_size=5,
                    name ="FHB QTL haplotypes",
                    annotations = list(
                      'DIS'=list(
                        aes=aes(x=intersection, y=DIS),
                        geom=list(
                          geom_quasirandom(aes(color=FIRST_YEAR), na.rm=TRUE,show.legend=F, size=1, alpha=0.6),
                          geom_boxplot(fill=NA, na.rm=TRUE, width=0.3, color="maroon",linewidth=1,outliers = F),
                          scale_color_viridis_c(name="First Year", limits=c(1995,2024))
                        )
                      ),
                      'VSK'=list(
                        aes=aes(x=intersection, y=VSK),
                        geom=list(
                          geom_quasirandom(aes(color=FIRST_YEAR), na.rm=TRUE, size=1, alpha=0.6),
                          geom_boxplot(fill=NA, na.rm=TRUE, width=0.3, color="maroon",linewidth=1,outliers = F),
                          scale_color_viridis_c(name="First Year", limits=c(1995,2024))
                        )
                      ),
                      'DON'=list(
                        aes=aes(x=intersection, y=DON),
                        geom=list(
                          geom_quasirandom(aes(color=FIRST_YEAR), na.rm=TRUE,show.legend=F, size=1, alpha=0.6),
                          geom_boxplot(fill=NA, na.rm=TRUE, width=0.3, color="maroon",linewidth=1,outliers = F),
                          scale_color_viridis_c(name="First Year", limits=c(1995,2024))
                        )
                      )
                      
                    )
                    
)
ggsave(here("output","figures",paste0(pop_code,"_FHB_3QTLs_effect_3traits_upset_v3.png")),
       height=10,width=8,scale=0.9)
```





# Appendix

```{r info}
t1 <- proc.time(); t1 - t0
print(sessionInfo(), locale=FALSE)
```


