---
title: "Create inputs for FHB_GGAIN project"
author: Charlotte Brault
output: html_document
date: '`r format(Sys.time(), "%d %B, %Y")`'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Libraries

```{r load libraries}
suppressPackageStartupMessages(library(here)) # root directory
suppressPackageStartupMessages(library(data.table)) # load data
suppressPackageStartupMessages(library(apercu)) # ap
suppressPackageStartupMessages(library(dplyr)) # formatting
suppressPackageStartupMessages(library(purrr)) # list handling
suppressPackageStartupMessages(library(ggvenn)) # Venn diagram
suppressPackageStartupMessages(library(bigstatsr)) # theme
suppressPackageStartupMessages(library(stringr)) # string handling
suppressPackageStartupMessages(library(janitor)) # formatting
suppressPackageStartupMessages(library(forcats)) # order factors
suppressPackageStartupMessages(library(ggh4x)) # nice plot
suppressPackageStartupMessages(library(gt)) # nice tables
suppressPackageStartupMessages(library(ggridges)) # ridge plot

```
This R chunk is used to assess how much time it takes to execute the R code in this document until the end:



# Set up root directory

```{r current dir}
here::i_am("analysis/explore_inputs.Rmd")
source(here("code","useful_functions.R"))
```


This R chunk is used to assess how much time it takes to execute the R code in this document until the end:

```{r time_0}
t0 <- proc.time()
```


# Load inputs

```{r}
input <- readRDS(here("data","all_inputs_URN_URSN_URNFHB.rds"))
ap(input)
map(input, \(x) map_dfr(x,n_distinct)) 

# ## remove Fhb5AS prior to 2007
# input$KASP <- input$KASP |> 
#   filter(!(QTL.name %in% "Fhb5AS" & FIRST_YEAR < 2007))
dat <- input$ALL
dat.fhb <- fread(here("data","dat_FHB_URSN_URNFHB.tsv"))
traits.ursn <- c("DIS","VSK","DON")
dat.info <- input$INFO
map_dbl(dat.info, n_distinct)
dat.ursn <- input$URSN
dat.urn.fhb <- filter(input$URN.FHB, YEAR >= 1995)

gid.ursn <- unique(input$URSN$GID)
gid.urn <- unique(input$URN$GID)
gid.urn.fhb <- unique(input$URN.FHB$GID)
gid.all <- unique(dat$GID)
kasp.gid <- unique(input$KASP$GID)
length(gid.all)
setdiff(input$URSN$LOC, input$URN$LOC) # 1 location in URSN not in URN FHB
all(kasp.gid %in% gid.all)
length(intersect(kasp.gid, gid.all))
length(intersect(kasp.gid, input$INFO$GID)) # 1545
```

Key numbers:

* URSN: 817 GID, 11 LOC, 30 YEARS

* URN: 1035 GID, 44 LOC, 57 YEARS

* URN FHB: 701 GID, 19 LOC, 31 YEARS

* ALL: 1758 GID, 45 LOC, 57 YEARS

# Explore data


## Experimental design

Plot a 3-panels heatmap of the number of genotypes tested in each location-year combination for the URN, URN FHB and URSN populations.

```{r exp_design}
# dat.exp <- rbind(dat.urn[,c("GID","LOC","YEAR","NURSERY")],
#                  dat.urn.fhb[,c("GID","LOC","YEAR","NURSERY")],
#                  dat.ursn[,c("GID","LOC","YEAR","NURSERY")])
dat.exp <- dat |> dplyr::select(GID, LOC, YEAR, NURSERY) 
loc.info <- fread(here("data","URN_trial_info.csv")) |> rename(LOC=name_short)
#dat.exp <- dat.exp |> filter(YEAR >= 1995)

dat.exp |> group_by(LOC,YEAR,NURSERY) |> 
  summarize(n=n(),.groups="drop") |> 
  mutate(LOC=as.factor(LOC)) |> 
  left_join(loc.info) |> 
  mutate(LOC.st=paste0(LOC,"&",state_short),
         state_short=as.factor(state_short),
         state_short=forcats::fct_infreq(state_short))|>
  arrange(state_short) |>  
  mutate(LOC.st=fct_inorder(LOC.st)) |>
  ggplot(aes(x=LOC.st,y=YEAR, fill=n))+
  facet_wrap(~NURSERY, ncol=1)+
  geom_tile() + 
  paletteer::scale_fill_paletteer_c("grDevices::ag_Sunset") +
  theme_bigstatsr(size.rel=0.8)+
  labs(x="Location", y="Year", fill="Number of genotypes")+
  guides(x = legendry::guide_axis_nested(),
         fill=guide_colorbar(barwidth=15,label.position="bottom"))+
  theme(axis.text.x = element_text(angle = 90, vjust =0.5),
        legend.direction="horizontal",legend.position="bottom",
        legend.title.position="top",legend.title = element_text(hjust=0.5,size=12),
        legend.text = element_text(size=10),
        strip.background = element_blank())
ggsave(here("output","figures","URN_URN-FHB_URSN_experimental_design_heatmap.png"), width=8, height=6)
```



## Distribution

```{r geom_density}

for(tr in traits.ursn){
  plot(ggplot(dat.fhb, aes(x=(.data[[tr]]), fill=NURSERY, color=NURSERY))+
         geom_density( alpha=0.5)+
         ggtitle(tr)+
         geom_rug()+ xlab(tr)+
         bigstatsr::theme_bigstatsr(size.rel=0.8)+
         paletteer::scale_colour_paletteer_d("colorblindr::OkabeIto_black") +
         paletteer::scale_fill_paletteer_d("colorblindr::OkabeIto_black") +
         theme(legend.position = "bottom"))
}
```

```{r ggridges_HD, fig.height=8, fig.width=5}
dat2plot <- dat.fhb |> 
  dplyr::select(YEAR, NURSERY, HD) |> distinct() |> 
  filter(!is.na(HD)) |> mutate(YEAR=as.factor(YEAR))
dat2plot$YEAR <- droplevels(as.factor(dat2plot$YEAR))
table(dat2plot$YEAR)
ggplot(data=dat2plot, aes(x=HD, y=YEAR, fill=NURSERY)) +
  geom_density_ridges(alpha=0.5) +
  theme_bigstatsr(size.rel=0.8)
```



```{r geom_points}
for(tr in traits.ursn){
  plot(ggplot(dat.fhb, aes(x=GID, y=.data[[tr]],color=LOC))+
         geom_point(size=1, alpha=0.5)+
         facet_wrap(~NURSERY)+
         ggtitle(tr)+
         theme_bigstatsr(size.rel=0.8)+
         guides(color=guide_legend(override.aes = list(size=3,alpha=1), ncol=8))+
         scale_color_manual(values=pal_loc()[dat.fhb$LOC])+
         theme(legend.position = "bottom",
               axis.text.x = element_blank(),
               panel.grid.major.x = element_blank()))
}
```


## Genotype composition

```{r}
## detect the genotypes tested in both nurseries
gid.both <- intersect(gid.ursn, gid.urn)
dat.info |> group_by(NURSERY,FIRST_YEAR) |> 
  summarise(n=n()) |> filter(FIRST_YEAR >= 1995) |> 
  ggplot(aes(x=FIRST_YEAR, y=n, fill=NURSERY))+
  geom_col(color="white")+ ylab("Number of genotypes") + xlab("First Year")+
  theme_bigstatsr(size.rel=0.8)+
  scale_x_continuous(breaks=seq(1995,2024,5))+
  paletteer::scale_fill_paletteer_d("colorblindr::OkabeIto_black", name="Nursery") +
  theme(panel.grid = element_blank(),legend.position = "bottom")

ggsave(here("output","figures","URN_URSN_genotype_composition_firstYear.png"), width=8, height=6)

```
Genotype composition by source for each nursery, including genotypes tested in both nurseries
```{r}
pal <- c("AAFC-CA"="#F5586A",MN="#ffcc33",ND="#2E8B57",SD="#3A5FCD",
         MT="#bd9b04",PRIVATE="#8A42C9",OTHER="grey30")
geno.compo <- rbind(data.frame(GID=gid.ursn, NURSERY="URSN"),
                    data.frame(GID=gid.urn, NURSERY="URN"))
geno.compo <- merge(geno.compo, dat.info[,c("GID","SOURCE","FIRST_YEAR")],
                    by="GID", all.x=TRUE)
geno.compo |> group_by(SOURCE,NURSERY,FIRST_YEAR) |> 
  summarise(n=n(),.groups="drop") |>
  ggplot(aes(x=FIRST_YEAR, y=n, fill=SOURCE))+
  facet_wrap(~NURSERY)+
  geom_col(color="white")+ ylab("Number of genotypes") + xlab("First Year")+
  theme_bigstatsr(size.rel=0.8)+
  scale_x_continuous(breaks=seq(1995,2024,5))+
  scale_fill_manual(values=pal, name="Source")+
  #paletteer::scale_fill_paletteer_d("colorblindr::OkabeIto_black", name="Nursery") +
  theme(panel.grid = element_blank(),legend.position = "bottom")
ggsave(here("output","figures","URSN_URN_genotype_composition_source_1995-2024.png"),
       width=8, height=5, scale=0.9)
tabyl(geno.compo, NURSERY, SOURCE)

table(geno.compo$SOURCE)
```


## Phenotypic correlation

### Between nurseries

```{r}
corr.nurseries <- c()
for(tr in traits.ursn){
  
  dat.fhb |>  dplyr::summarise(n = dplyr::n(), .by = c(GID, LOC, YEAR, ENV, NURSERY)) |>
  dplyr::filter(n > 1L)
  dat2plot <- dat.fhb |> dplyr::select(GID, LOC, YEAR, ENV, NURSERY, tr) |>
    tidyr::pivot_wider(names_from = NURSERY, values_from = tr) |> 
    filter(!is.na(URSN) | !is.na(URN_FHB))
  
  summary(dat2plot$URSN)
  corr.nurseries[tr] <- round(cor(dat2plot$URSN, dat2plot$URN_FHB, use="pairwise.complete.obs"),3)
  plot(ggplot(dat2plot, aes(x=URSN, y=URN_FHB))+
         geom_point()+
         geom_abline(slope=1, intercept=0, linetype="dashed", color="grey30")+
         ggtitle(tr)+
         theme_bigstatsr(size.rel=0.8)+
         theme(legend.position = "bottom"))
}
corr.nurseries
```

### Between traits

```{r}
GGally::ggcorr(dat.ursn[,traits.ursn], label=TRUE, label_size=5, geom="tile",max_size=11,
               low="#FF3D7FFF", mid="#E0E4CCFF",high="#69D2E7FF",
               size=6,label_round=2)+
  ggtitle("URSN")
GGally::ggcorr(dat.urn.fhb[,traits.ursn], label=TRUE, label_size=5, geom="tile",max_size=11,
               low="#FF3D7FFF", mid="#E0E4CCFF",high="#69D2E7FF",
               size=6,label_round=2)+
  ggtitle("URN FHB")
```

```{r}
ggplot(dat.ursn, aes(x=DIS, y=DON))+
  geom_point(alpha=0.5)+
  geom_smooth(method="lm", color="blue")+
  theme_bigstatsr(size.rel=0.8)+
  ggtitle("URSN")+
  xlab("Disease index (DIS)")+
  ylab("Deoxynivalenol (DON) content (ppm)")
dat.ursn |> filter(DIS < 15, DON > 15) |> 
  select(GID, ENV, DIS, DON, VSK) |> 
  print_table()

```


## Key numbers for publication

```{r}
purrr::map_dbl(dat.fhb, n_distinct)
purrr::map_dbl(dat.ursn, n_distinct)
purrr::map_dbl(dat.urn.fhb, n_distinct)
length(intersect(unique(dat.ursn$GID), unique(dat.urn.fhb$GID))) ## 125
length(intersect(unique(dat.ursn$LOC), unique(dat.urn.fhb$LOC))) ## 8
length(intersect(unique(dat.ursn$YEAR), unique(dat.urn.fhb$YEAR))) ## 28
dat.fhb |> group_by(NURSERY, ENV) |> summarize(n=n()) |> 
  group_by(NURSERY) |> summarize(n_mean=mean(n),
                                 n_min=min(n),
                                 n_max=max(n))
```
Check if the URSN checks (Wheaton, BacUP and ND2710) were constantly present in the URN FHB population:
```{r}
check.FHB.gid <- c("WHEATON","BACUP","ND2710")
check.urn.gid <- c("CHRIS","MARQUIS")
urn.fhb.checks <- dat.urn.fhb |> select(GID, LOC,YEAR,ENV) |> 
  filter(GID %in% check.FHB.gid) |> distinct()
length(setdiff(unique(dat.urn.fhb$ENV),unique(urn.fhb.checks$ENV)))
dat.urn.fhb |> group_by(GID) |> summarize(n=n()) |> arrange(desc(n)) |> head(n=10)

urn.fhb.checks <- dat.urn.fhb |> select(GID, LOC,YEAR,ENV) |> 
  filter(GID %in% check.urn.gid) |> distinct()
length(setdiff(unique(dat.urn.fhb$ENV),unique(urn.fhb.checks$ENV)))
```



# Build table of fit information for URN and URSN

Data generated in `extract_genoVal_URN_URSN_FHB.Rmd` and `extract_genoVal_URN.Rmd` scripts.

TODO: add crop ontology codes

```{r}
fit.agro <- fread(here("output","URN_lme4_fit-info_all_traits.tsv")) |> 
  mutate(pop="URN Agro", .before=trait) |> 
  filter(trait %in% c("YLD","PROT","TWT"))
fit.fhb <- fread(here("output","URN_URSN_FHB_lme4_fit-info_all_traits.tsv")) |> 
  mutate(pop="URN FHB & URSN", .before=trait) |> filter(trait %in% c("DIS","VSK","DON"))


fit.info <- rbind(fit.agro, fit.fhb)


gt.format <- fit.info |>
  gt(groupname_col="pop",
     row_group_as_column = TRUE) |> 
  cols_align("center") |> 
  # tab_header(
  #   title = "Fit information for URN and URSN populations",
  #   subtitle = "LME4 model fit information for all traits"
  # ) |> 
  cols_label(
    trait = "Trait",
    pop = "Population",
    transformation = "Transformation",
    missing.data.percent = "Missing data (%)",
    fixed.effects = "Fixed effects",
    random.effects = "Random effects",
    min = "Min.", max="Max.", mean="Mean",sd="Standard deviation",
    reliability = "Reliability",
    ngeno = "Number of genotypes",
    nobs = "Number of observations"
  ) |> 
  fmt_number(columns=2:8, decimals=0) |> 
  fmt_number(columns=9, decimals=2) |> 
  # tab_options(row_group.padding=px(4),
  #             data_row.padding = px(2),
  #   summary_row.padding = px(3)) |> 
  opt_stylize(style = 5, color = 'gray')
gt.format
gtsave(gt.format, here("output","figures","table_URN_URSN_lme4_fit-info_all_traits.png"),
        zoom=2, expand=5)

gtsave(gt.format, here("output","figures","table_URN_URSN_lme4_fit-info_all_traits.tex"))
```



# Appendix

```{r info}
t1 <- proc.time(); t1 - t0
print(sessionInfo(), locale=FALSE)
```

