---
title: "Extract genotypic values for URN Agro phenotypic data"
author: Charlotte Brault, adapted from Timothée Flutre
output: html_document
date: '`r format(Sys.time(), "%d %B, %Y")`'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Libraries

```{r load libraries}
suppressPackageStartupMessages(library(here)) # root directory
suppressPackageStartupMessages(library(data.table)) # load data
suppressPackageStartupMessages(library(lattice)) # plot
suppressPackageStartupMessages(library(apercu)) # ap
suppressPackageStartupMessages(library(dplyr)) # formatting
suppressPackageStartupMessages(library(knitr)) # table display
suppressPackageStartupMessages(library(rmarkdown)) # table display
suppressPackageStartupMessages(library(gtsummary)) # data summary
suppressPackageStartupMessages(library(ggplot2)) # plot
suppressPackageStartupMessages(library(GGally)) # pairsplot & parallel
suppressPackageStartupMessages(library(paletteer)) # plot palette
suppressPackageStartupMessages(library(janitor)) # formating
suppressPackageStartupMessages(library(tidyr)) # pivot
suppressPackageStartupMessages(library(crosstable)) # crosstable
suppressPackageStartupMessages(library(stringr)) # string
suppressPackageStartupMessages(library(stringdist)) # string distance
suppressPackageStartupMessages(library(rutilstimflutre)) # LMM devtools timflutre
suppressPackageStartupMessages(library(lmerTest))
suppressPackageStartupMessages(library(boot)) #bootstraping
suppressPackageStartupMessages(library(bigstatsr)) # plot theme

```


# Set up root directory

```{r current dir}
here::i_am("analysis/extract_genoVal_URN.Rmd")
source(here("code","useful_functions.R"))
```


This R chunk is used to assess how much time it takes to execute the R code in this document until the end:

```{r time_0}
t0 <- proc.time()
par <- "URN" ## optional parameter for fitting, for model comparison
```


# Load phenotypic data


Data was curated in `create_inputs_explore.Rmd`. [See report](docs/create_inputs_explore.html)


```{r load_phenotypic_data, warning=FALSE}
p2f <- here("data","all_inputs_URN_URSN_URNFHB.rds")
file.mtime(p2f)
all.inputs <- readRDS(p2f)
ap(all.inputs)
dat = all.inputs$URN
dim(dat) ## 29714 x 12
head(dat)
(traits <- sapply(dat, class) %>% {.[.=="numeric"]} %>% names())
info.cols <- setdiff(colnames(dat),traits)
dat <- dat %>% mutate_at(all_of(info.cols), as.factor) |> as.data.frame()
#dat <- rename(dat, "GID"="line","SW30"="30SSW")
(miss.dat.perc <- round(100* (nrow(dat) -apply(dat, 2,function(x) sum(!is.na(x)))) /nrow(dat)))
summ.traits <- as.data.frame(apply(t(apply(dat[,traits],2,summary)),2,round, 1))
summ.traits$SD <- round(apply(dat[,traits],2,sd,na.rm=TRUE),1)
summ.traits$traits <- rownames(summ.traits)
print_table(as.data.frame(summ.traits))
```

Apply transformation 
```{r}
#transf <- c(LODG="log10")
## apply transformation to ensure normality and homoscedasticity
#dat <- dat |> mutate(LODG = log10(LODG))
transf <- c(rep(NA,length(traits))) ; names(transf) <- traits
for(tr in traits){
  plot(ggplot(dat, aes(x=(.data[[tr]])))+
         geom_density( alpha=0.5, fill="grey50")+
         ggtitle(tr)+ geom_rug()+ xlab(tr)+
         bigstatsr::theme_bigstatsr(size.rel=0.8))

}
```
Update genotype names
```{r}
purrr::map_dbl(dat, n_distinct)

```



# Model fitting, comparison and selection

## Fitting the global model

```{r model_fit_comp_sel_fixed_geno, warning=FALSE}
out <- list()
dat.noNA <- list()
dat <- as.data.frame(dat)
for(tr in traits){
  write(paste0(tr,"\n"),stderr())
  dat.noNA[[tr]] <- na.omit(droplevels(dat[,c(tr, "GID","YEAR","LOC")]))
  
  if(length(unique(as.character(dat.noNA[[tr]]$YEAR))) < 2){
    next
  } else {
    glob.form <- paste0(tr, " ~ 1",
                        " + GID ",
                        " + YEAR ",
                        " + (1|LOC) ",
                        " + (1|YEAR:LOC)"

                        
    )
    
    out[[tr]] <- plantTrialLmmFitCompSel(glob.form=glob.form,
                                         dat=dat.noNA[[tr]],
                                         part.comp.sel=c("random","fixed"),
                                         nb.cores=1, verbose=1,
                                         saved.file = here("output","lme4",paste0("lme4_save_fit_",tr,"_",par, ".RData"))
    )
    
  }
}
```


Problematic traits: those with one YEAR of phenotyping (no rep to estimate BLUP)

```{r}
(pbm <- names(which(unlist(lapply(out, class)) == "try-error")))
```

Amount of missing data:
```{r miss_dat}
dat <- janitor::remove_empty(dat,which=c("rows", "cols"))

(miss.dat.perc <- round(100 * (nrow(dat)- unlist(lapply(dat.noNA,nrow))) / nrow(dat),1))
ngeno <- unlist(lapply(dat.noNA,function(x) length(unique(x$GID))))
```


## Handle case where "geno" is not kept in the final model

```{r refit}
for(tr in traits[!traits %in% pbm]){
  if(! "GID" %in% out[[tr]]$best.preds.fix){
    print(tr)
    glob.form <- paste0(tr, " ~ 1",
                        " + GID",
                        " + YEAR",
                        " + (1|LOC) ",
                        " + (1|YEAR:LOC)"

    )
    
    out[[tr]] <- try(plantTrialLmmFitCompSel(glob.form=glob.form,
                                             dat=dat.noNA[[tr]],
                                             part.comp.sel=c("fixed"),
                                             nb.cores=1, verbose=0))
  }
}
```



## Analyze variance components

```{r}
for(tr in traits[!traits %in% pbm]){
  cat(paste(tr,"\t"))
  try(print(anova(out[[tr]]$bestmod.reml)))
  try(print(ranova(out[[tr]]$bestmod.ml)))
  #anov <- rand(out[[tr]]$bestmod.reml)
}
```


# Assumption checking

## Residual preparation and blues prediction

```{r}
geno.blues <- list() ; geno.var.blues <- list()
fit.all <- list()
for(tr in traits[!traits %in% pbm]){
  fit.all[[tr]] <- cbind(out[[tr]]$dat,
                         response=out[[tr]]$dat[[tr]],
                         cond.res=NA,
                         scl.cond.res=NA,
                         fitted=NA)
  fit.all[[tr]]$cond.res <- residuals(out[[tr]]$bestmod.reml)
  fit.all[[tr]]$scl.cond.res<- residuals(out[[tr]]$bestmod.reml) /
    sigma(out[[tr]]$bestmod.reml)
  fit.all[[tr]]$fitted <- fitted(out[[tr]]$bestmod.reml)
  fixEf <- fixef(out[[tr]]$bestmod.reml)[-1]
  geno.blues[[tr]] <- fixEf[grep(names(fixEf), pattern="GID")]
  geno.blues[[tr]] <- fixef(out[[tr]]$bestmod.reml)["(Intercept)"] + geno.blues[[tr]]
  names(geno.blues[[tr]]) <- gsub("GID","",names(geno.blues[[tr]]))
}
```


## Error homoscedasticity

Overall
```{r check_error_homoscedasticity}
for(tr in traits[!traits %in% pbm]){
  x.lim <- max(abs(fit.all[[tr]]$scl.cond.res), na.rm=TRUE)
  plot(x=fit.all[[tr]]$scl.cond.res, y=fit.all[[tr]]$fitted, las=1,
       xlim=c(-x.lim, x.lim),
       xlab="scaled conditional residuals",
       ylab="fitted responses",
       main=tr)
  abline(v=c(0, -1.96, 1.96), lty=c(2, 3, 3))
}
```

Color per YEAR

```{r}
for(tr in traits[!traits %in% pbm]){
  fit.all[[tr]]$YEAR <- as.factor(fit.all[[tr]]$YEAR)
  x.lim <- max(abs(fit.all[[tr]]$scl.cond.res), na.rm=TRUE)
  g <- ggplot(fit.all[[tr]], aes(x=scl.cond.res, y=fitted,color=YEAR)) +
    geom_point() + theme_bw() + 
    xlim(c(-x.lim, x.lim)) +
    xlab("scaled conditional residuals") +
    ylab("fitted responses") +
    ggtitle(tr) +
    paletteer::scale_colour_paletteer_d("Polychrome::palette36") +
    geom_vline(xintercept = c(0, -1.96, 1.96),
               linetype="dashed")
  plot(g)
  
}
```


## Error normality

```{r check_error_normality}
for(tr in traits[!traits %in% pbm]){
  shapiro.test(sample(fit.all[[tr]]$scl.cond.res,5000, replace=T))
  qqnorm(y=fit.all[[tr]]$scl.cond.res,
         main=paste0(tr, ": scaled conditional residuals"))
  qqline(y=fit.all[[tr]]$scl.cond.res, col="red")
}
```


## Outlying genotypes

Print genotypic values and extreme values:
```{r check_outlying_genos_plot}
for(tr in traits[!traits %in% pbm]){
  
  #x.lim <- max(abs(geno.blues[[tr]]))
  par(mar=c(5,6,4,1))
  plot(x=geno.blues[[tr]], y=1:length(geno.blues[[tr]]),
       #xlim=c(-x.lim, x.lim),
       xlab="genotypic blues",
       main=paste0(tr),
       pch=19,
       yaxt="n", ylab="")
  axis(side=2, at=1:length(geno.blues[[tr]]), labels=names(geno.blues[[tr]]), las=1)
  #abline(v=0, lty=2)
  idx <- which.max(geno.blues[[tr]])
  text(x=geno.blues[[tr]][idx], y=idx, labels=names(idx), pos=2)
  idx <- which.min(geno.blues[[tr]])
  text(x=geno.blues[[tr]][idx], y=idx, labels=names(idx), pos=4)
}
```

## Normality of genotypic blues

```{r check_normality_geno_blues_qqplot}
for(tr in traits[!traits %in% pbm]){
  shapiro.test(geno.blues[[tr]])
  qqnorm(y=geno.blues[[tr]], asp=1,
         main=paste0(tr, ": genotypic blues"))
  qqline(y=geno.blues[[tr]], col="red")
}
```

## Independence between errors and genotypes

Color with location
```{r check_indep_error_geno_plot}
for(tr in traits[!traits %in% pbm]){
  x.lim <- max(abs(fit.all[[tr]]$scl.cond.res))
  plot(lattice::dotplot(GID ~ scl.cond.res, data=fit.all[[tr]],
                        xlim=c(-x.lim, x.lim),
                        groups=LOC,
                        pch=19,
                        xlab="scaled conditional residuals",
                        main=tr,
                        scales = list(y = list(draw=FALSE)),
                        panel=function(x,y,...){
                          panel.abline(v=c(0, -1.96, 1.96), lty=c(2,3,3))
                          panel.dotplot(x,y,...)
                        }))
}
```



```{r check_indep_error_geno_plot_YEAR_block}
for(tr in traits[!traits %in% pbm]){
  print(
    lattice::dotplot(GID ~ scl.cond.res | YEAR,
                     data=fit.all[[tr]],
                     groups=LOC,
                     pch=19,
                     auto.key=list(space="right"),
                     xlab="scaled conditional residuals",
                     main=tr,
                     scales = list(y = list(draw=FALSE)),
                     panel=function(x,y,...){
                       panel.abline(v=c(0, -1.96, 1.96), lty=c(2,3,3))
                       panel.dotplot(x,y,...)
                     }))
}
```



# Model outputs

## Summary

```{r bestmod_infer}
vc.reml  <- list()
for(tr in traits[!traits %in% pbm]){
  ## summary(bestmod.ml)
  (vc.ml <- as.data.frame(VarCorr(out[[tr]]$bestmod.ml)))
  #summary(bestmod.reml)
  vc.reml[[tr]] <- as.data.frame(VarCorr(out[[tr]]$bestmod.reml))
  print(vc.reml[[tr]])
}
```


## Broad-sense heritability 

```{r H2}
# H2_list <- list() ; mySumm <- list()
# for(tr in traits){
#   cat(tr)
#   H2_list[[tr]] <- estimH2means(dat=droplevels(dat.noNA[[tr]]), colname.resp=tr,
#                                 colname.trial="YEAR", vc=vc.reml[[tr]],
#                                 geno.var.blues=geno.var.blues[[tr]])
#   
#   print(H2_list[[tr]]$mean.nb.trials)
#   print(H2_list[[tr]]$mean.nb.reps.per.trial)
#   print(H2_list[[tr]]$H2.classic)
#   mySumm[[tr]] <- H2_list[[tr]]$sryStat
# }
```

## Reliability

```{r}
rel <- c()
for(tr in traits){
  r <- psych::harmonic.mean(as.vector(table(dat.noNA[[tr]]$GID)), zero=FALSE)
  #vg <- vc.reml[[tr]]$vcov[vc.reml[[tr]]$grp == "GID"]
  vg <- var(geno.blues[[tr]])
  ve <- vc.reml[[tr]]$vcov[vc.reml[[tr]]$grp == "Residual"]
  rel[tr] <- round(vg/(vg+ve/r),3)
}
rel
```


# Save outputs

## Fit information

```{r info_to_save}

info2save <- data.frame(trait=traits,
                        transformation=c(transf[traits]),
                        missing.data.percent=round(miss.dat.perc[traits],0),
                        ngeno=c(ngeno[traits]),
                        nobs=sapply(dat.noNA,nrow),
                        #transformation=ifelse(transf == "id", "NA", transf),
                        fixed.effects=unlist(lapply(out, function(x)
                          paste0(x$best.preds.fix,collapse=", "))),
                        random.effects=unlist(lapply(out, function(x)
                          paste0(x$best.preds.rnd,collapse=", "))),
                        min=summ.traits[traits,"Min."],
                        max=summ.traits[traits,"Max."],
                        mean=summ.traits[traits,"Median"],
                        sd=summ.traits[traits,"SD"],
                        reliability=rel)

info2save

p2f <- here("output",paste0(par,"_lme4_fit-info_all_traits.tsv"))
write.table(x=info2save,
            file=p2f, quote=FALSE, sep="\t", row.names=FALSE)
tools::md5sum(path.expand(p2f))
file.mtime(p2f)
```



## Genotypic values


Format genotypic values
```{r}
geno_blues <- list()
for(tr in traits[!traits %in% pbm]){
  # Add blues for GID and cross
  geno_blues[[tr]] <-  data.frame(GID=names(geno.blues[[tr]]), 
                                  geno.blues=geno.blues[[tr]],
                                  trait=tr,
                                  stringsAsFactors=FALSE)
}
## de transform data
geno_blues$LODG$geno.blues <- 10^(geno_blues$LODG$geno.blues)

```



Plot genotypic values distribution
```{r}
for(tr in traits[!traits %in% pbm]){
  ## plot distribution per cross
  plot(ggplot(geno_blues[[tr]], aes(x=geno.blues)) +
         geom_density() +
         geom_rug() +
         theme_bw() +
         ggtitle(paste0("blues for ", tr," trait")))
  
}
ap(geno_blues)
```

Save genotypic values
```{r save_geno_values, echo=TRUE,warning=FALSE, message=FALSE}


# # save list of blues and associated variance
# p2f <- here("output",paste0("URSN_geno_blues_var_lme4-values",par,".Rdata"))
# save(geno_blues, file=p2f)
# tools::md5sum(path.expand(p2f))
# file.mtime(p2f) 

## save data frame of blues for all traits
geno_blues_long <- do.call(rbind, geno_blues)
head(geno_blues_long)
geno_blues_long$geno.var.blues <- NULL
df_blues <- pivot_wider(geno_blues_long,names_from  = c("trait"),
                        values_from = "geno.blues")
df_blues <- df_blues %>% mutate_if(is.numeric, round, 4)
dim(df_blues)
ap(df_blues)


# ## add first YEAR of trial
# df_blues <- merge(df_blues, df_YEAR, by.x="id_geno", by.y="GID", all.x=TRUE)
# df_blues <- relocate(df_blues, firstYEAR, .after = id_geno)
# head(df_blues)
# df_blues <- arrange(df_blues,firstYEAR)
p2f <- here("output",paste0(par,"_geno_blues_lme4-values.tsv"))
write.table(df_blues, file=p2f,col.names=TRUE, row.names=FALSE, sep="\t",
            fileEncoding = "UTF-8")
tools::md5sum(path.expand(p2f))
file.mtime(p2f) 
```

What to do with BLUEs with negative values?



# Plot correlations

```{r plot_corr, fig.height=8, fig.width=8}
# ggpairs(dplyr::select(Y,-GID),progress = FALSE) +
#   theme_bw()
ggcorr(df_blues,label=TRUE, 
       label_size=5, geom="tile",max_size=11,
       low="#FF3D7FFF", mid="#E0E4CCFF",high="#69D2E7FF", size=6,label_round=2)
ggsave(here("output","figures",paste0(par,"_GenoCorr_",length(traits),"traits.png")))
```

Next step: Complete the genotype information in `explo_geno_URN_URSN.Rmd` and study fit info and blues in `analyze_blues_URN_URSN.Rmd`.

# Appendix

```{r info}
t1 <- proc.time(); t1 - t0
print(sessionInfo(), locale=FALSE)
```



